---
title: "Loan Presentation"
author: "Si Tayeb Houari"
date: "2025-03-23"
output: html_document
---

#Overview:
Welcome to the Loan Default Prediction Challenge 2025! This competition is designed to help data scientists and machine learning enthusiasts practice their skills with a real-world financial dataset.
Your Goal: Your goal is to predict whether a loan applicant is at risk of defaulting based on their financial and personal information.

#Evaluation Metric
Submissions will be evaluated using the F1 Score, a metric that balances precision and recall:
F1=2× Precision+Recall/Precision×Recall


#1. Business Context & Real-World Impact
.Why does this problem matter?

.Who benefits from this model? (Banks, financial institutions, insurance companies?)

.How does accurate classification improve decision-making? (E.g., reducing loan defaults, improving financial stability)

.What are the consequences of misclassification?

.False Positives (predicting high risk when the person is actually low risk) → Lost business opportunities.

.False Negatives (predicting low risk when the person is actually high risk) → Increased loan defaults.

#2. Data Challenges & Solutions
.Imbalanced Data: Most real-world classification problems suffer from class imbalance. We addressed this using SMOTE (Synthetic Minority Oversampling Technique).

.Feature Selection: Some features may not be useful (e.g., Id or highly correlated variables).

.Missing Data: If applicable, how did you handle missing values? (e.g., imputation, removal)

# 3. Model Selection Justification
Explain why you chose Random Forest over other models:

✅ Why not Logistic Regression?

Too simple and assumes a linear decision boundary, which may not capture complex patterns in the data.

✅ Why not XGBoost?

While XGBoost is powerful, it may overfit when hyperparameters are not well-tuned.


✅ Why Random Forest?

Random Forest performed better in terms of accuracy and generalization in this dataset.

Random Forest is more robust to noisy data and does not require extensive hyperparameter tuning.

It provides strong performance with minimal tuning.

Works well with both categorical and numerical data.

Less sensitive to hyperparameter choices compared to boosting models.

Reduces variance and prevents overfitting by averaging multiple trees.
```{r}
# ✅ Load Required Libraries
library(tidymodels)
library(tidyverse)
library(skimr)
library(themis)  # Handling class imbalance
library(ggcorrplot)  # Correlation matrix visualization
library(ranger)  # Fast Random Forest
```




```{r}
# Load datasets
train <- read.csv(file.choose())
test <- read.csv(file.choose())

# Convert target variable to factor
train <- train %>%
  mutate(Risk_Flag = as.factor(Risk_Flag))  # Ensures classification

```


```{r}

head(train)
```




```{r}
skim(train)  # Summary statistics

```

#eda
```{r}
#Visualize Class Distribution"epresents the number of samples for each class (Risk_Flag)"

# Bar plot
ggplot(train, aes(x = as.factor(Risk_Flag))) + 
  geom_bar(fill = "steelblue") + 
  labs(title = "Class Distribution")

# Pie chart
train %>% 
  count(Risk_Flag) %>% 
  ggplot(aes(x = "", y = n, fill = as.factor(Risk_Flag))) + 
  geom_bar(stat = "identity", width = 1) + 
  coord_polar(theta = "y") + 
  labs(title = "Class Distribution (Pie Chart)", fill = "Risk_Flag") + 
  theme_void()
#class proportions 
train %>%
  count(Risk_Flag) %>%
  mutate(prop = n / sum(n))
```



```{r}
#Numeric Variable Distributions
train %>%
  pivot_longer(cols = c(Income, Age, Experience, CURRENT_JOB_YRS, CURRENT_HOUSE_YRS)) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "green", alpha = 0.7) +
  facet_wrap(~name, scales = "free") +
  labs(title = "Distribution of Numeric Variables")

```
2️⃣ Why Is This Important?
Helps identify skewness (e.g., right-skewed income distribution).
Detects outliers (e.g., extreme values in income or experience).
Shows data spread, indicating whether variables need transformation (e.g., log transformation for highly skewed distributions).

3️⃣ Key Takeaways
Income: Often right-skewed; may require log transformation.
Age & Experience: Should follow a reasonable distribution. Large gaps or outliers might indicate data issues.
CURRENT_JOB_YRS & CURRENT_HOUSE_YRS: Shorter durations could indicate frequent job or house changes, which may be useful for prediction.

##

Key Observations from the Distribution Plots
1️⃣ Age:

The distribution appears uniform, except for a peak around age 20.

The presence of a spike suggests a potential data imbalance or a large group of young applicants.

2️⃣ Experience:

The data is fairly evenly distributed.

This suggests that experience is well-spread across different levels, which is good for modeling.

3️⃣ Income:

The distribution is approximately uniform, but with a concentration towards lower values.

Suggests a wide income range, but many applicants earn lower salaries.

4️⃣ Current Job Years:

The distribution is right-skewed, meaning most applicants have been at their job for a short period.

Longer tenure is less frequent, which could impact loan stability and default risk.

5️⃣ Current House Years:

The distribution has distinct peaks at specific values (e.g., 10, 12, 13, 14 years).

Indicates data irregularities or possible data collection issues (e.g., rounding or predefined categories).

Further investigation is needed—perhaps this feature was recorded in buckets rather than as continuous data.

```{r}
#Relationship Between Variables and Risk_Flag

ggplot(train, aes(x = Risk_Flag, y = Income)) +
  geom_boxplot(fill = "orange", alpha = 0.6) +
  labs(title = "Income vs Loan Default Risk", x = "Risk_Flag", y = "Income")

ggplot(train, aes(x = Risk_Flag, y = Age)) +
  geom_boxplot(fill = "purple", alpha = 0.6) +
  labs(title = "Age vs Loan Default Risk", x = "Risk_Flag", y = "Age")

```

1️⃣ Age vs Loan Default Risk
The first boxplot compares Age between customers who defaulted (Risk_Flag = 1) and those who did not (Risk_Flag = 0).

The median age is similar across both groups, indicating no strong age-related trend in default risk.

2️⃣ Income vs Loan Default Risk
The second boxplot compares Income across default and non-default groups.

The distribution appears nearly identical, suggesting that income alone does not significantly differentiate risk levels.

```{r}

#cor_matrix
cor_matrix <- train %>%
  select(where(is.numeric)) %>%
  cor()

ggcorrplot(cor_matrix, lab = TRUE, colors = c("red", "white", "green"))


```
#modeling


✔ income_per_exp → Measures how much income a person earns per year of experience.
✔ stability_index → Evaluates job stability based on experience.
✔ own_both → Identifies individuals with strong asset ownership.

These features add valuable insights for predicting loan risk, helping the model make better decisions



```{r}
# ✅ Load Required Libraries
library(tidymodels)
library(tidyverse)
library(skimr)
library(themis)  # Handling class imbalance
library(ggcorrplot)  # Correlation matrix visualization
library(ranger)  # Fast Random Forest

# ✅ Convert Target Variable to Factor
train <- train %>%
  mutate(Risk_Flag = as.factor(Risk_Flag))  # Ensures classification

# ✅ Feature Engineering: Create New Features
train <- train %>%
  mutate(
    income_per_exp = Income / (Experience + 1),  # Avoid division by zero
    stability_index = ifelse(Experience > 0, CURRENT_JOB_YRS / Experience, 0),
    own_both = ifelse(Car_Ownership == "Yes" & House_Ownership == "Owned", "Yes", "No")
  )

test <- test %>%
  mutate(
    income_per_exp = Income / (Experience + 1),
    stability_index = ifelse(Experience > 0, CURRENT_JOB_YRS / Experience, 0),
    own_both = ifelse(Car_Ownership == "Yes" & House_Ownership == "Owned", "Yes", "No")
  )

# ✅ Define Preprocessing Recipe
loan_recipe <- recipe(Risk_Flag ~ ., data = train) %>%
  update_role(Id, new_role = "ID") %>% # Id stays in the dataset but won’t be used as a predictor
  step_zv(all_predictors()) %>%  # Remove zero-variance predictors
  step_log(Income, base = 10) %>%  # Log-transform income
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%  # One-hot encoding
  step_normalize(all_numeric_predictors()) %>%  # Normalize numeric variables
  step_smote(Risk_Flag)  # Handle class imbalance

# ✅ Define Random Forest Model (Default Hyperparameters)
rf_model <- rand_forest(
  trees = 500  # Number of trees (default)
) %>%
  set_engine("ranger") %>%
  set_mode("classification")

# ✅ Create Workflow
rf_wf <- workflow() %>%
  add_model(rf_model) %>%
  add_recipe(loan_recipe)

# ✅ Fit Model on Training Data
rf_fit <- rf_wf %>%
  fit(data = train)

# ✅ Predict on Test Data
rf_preds <- predict(rf_fit, test, type = "prob") %>%
  bind_cols(predict(rf_fit, test, type = "class")) %>%
  bind_cols(test %>% select(Id))

# ✅ Convert Probabilities to Class Labels
rf_preds <- rf_preds %>%
  mutate(Risk_Flag = ifelse(.pred_1 >= 0.5, 1, 0)) %>%
  select(Id, Risk_Flag)

# ✅ Model Performance Evaluation on Training Data
train_preds <- predict(rf_fit, train, type = "prob") %>%
  bind_cols(predict(rf_fit, train, type = "class")) %>%
  bind_cols(train %>% select(Risk_Flag))

# Ensure factors for comparison
train_preds <- train_preds %>%
  mutate(.pred_class = as.factor(.pred_class),
         Risk_Flag = as.factor(Risk_Flag))

# Compute Metrics
train_metrics <- train_preds %>%
  metrics(truth = Risk_Flag, estimate = .pred_class)

# Compute F1-score
train_f1 <- train_preds %>%
  f_meas(truth = Risk_Flag, estimate = .pred_class)


```


```{r}

# Print Training Metrics
print("📊 Model Performance on Training Data:")
print(train_metrics)
print("🔹 F1-Score:")
print(train_f1)
```


```{r}
# ✅ Save Submission File
submission_path <- "C:/Users/Propriétaire/Downloads/submission_rf_fixed.csv"
write_csv(rf_preds, submission_path)

cat("✅ Submission file saved successfully at:", submission_path)

```




